---
title: "SPRAAKLAB validation study"
author: "Teja Rebernik et al."
date: "Generated on: `r date()`"
output: 
  html_document:
    toc: true
    code_folding: show
    toc_float: 
        collapsed: false
        smooth_scroll: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following script documents the statistical analysis used to assess and validate the use of SPRAAKLAB, a mobile laboratory. The current version was prepared for initial submission to journal Behavior Research Methods. 

# Data preparation 

```{r}
setwd("C:/Users/Teja/Desktop/SPRAAKLAB") #set your working directory

#load necessary packages
library(plotfunctions)
library(dplyr)
library(mgcv)
library(itsadug)
library(ggplot2)
library(ggpubr)

#load dataset
dat <- read.csv("spraaklab_dataset_public.csv", header=TRUE)

```

The dataset includes the following variables: 
*Participant: only participants who consented to having their preprocessed data shared with other researchers 
*Participant sex: factor with two levels, female and male
*Participant age
*Trial: consecutive trial of the experiment
*formantZ: z-transformed formant value (per formant, per participant: mean of all baseline trials subtracted from each trial, divided by the standard deviation of all baseline trials)
*FormantType: factor with two levels, F1 and F2
*Location: factor with two levels, lab (traditional laboratory) and spraaklab (mobile laboratory)

```{r}
#Inspect the data
head(dat)
str(dat)

#make sure the variables are factors
dat <- dat %>%
  mutate(across(c(participant, targetWord, location, FormantType, sex), as.factor))

#order the data

dat <- dat[order(dat$FormantType, dat$participant, dat$Trial), ] # sort data per trial

head(dat)
str(dat)

```
# Generalized Additive Modelling

For more information on model-fitting procedures we used, see Wieling (2018).

## Binary curves

```{r}
#make binary curves to compare locations, where SPRAAKLAB is 1 and lab is 0

dat$isSpraaklabF1 <- (dat$location == 'spraaklab' & dat$FormantType == "F1")*1
dat$isSpraaklabF2 <- (dat$location == 'spraaklab' & dat$FormantType == "F2")*1
```

##  GAMM

```{r}
location_model <- bam(formantZ ~ s(Trial, by=FormantType) + FormantType +
                        
                        s(Trial, by=isSpraaklabF1) + s(Trial, by=isSpraaklabF2) +
                        
                        s(Trial, participant, by=FormantType, bs='fs', m=1) + 
                        
                        s(Trial, targetWord, by=FormantType, bs='fs', m=1) +
                        
                        s(Trial, targetWord, by=isSpraaklabF1, bs='fs', m=1) + 
                        s(Trial, targetWord, by=isSpraaklabF2, bs='fs', m=1), 
                        
                      data=dat, discrete=T, nthreads=8, family='scat')

summary(location_model, re.test=F)
```
## Check residuals 

```{r}
qqnorm(resid(location_model))
qqline(resid(location_model))
acf_resid(location_model)
```

## Plot model output

```{r}
#plot both models in colourblind-friendly colours

turqoise <- c("#004D40")
yellow <- c("#FFC107")
blue <- c("#1E88E5")
pink <- c("#D81B60")
black <- c("#000000")

#plot F1
plot_smooth(location_model, view="Trial", cond=list(isSpraaklabF1 = 0, FormantType='F1'), rug = F, main = "F1 over trials", ylab = "F1 (z-transformed)", xlab = "Trials",xlim=c(1,114), ylim=c(-4,3), col=c(yellow), cex.lab=1.3, cex.axis = 1.3, hide.label = TRUE)
plot_smooth(location_model, view="Trial", cond=list(isSpraaklabF1 = 1, FormantType='F1'), rug = F, main = "F1 over trials", ylab = "F1 (z-transformed)", xlab = "Trials",xlim=c(1,114), ylim=c(-4,3), col=c(blue), add = T)
par(abline(h=0, v=c(1, 25,49,97, 114), lty=3)) #add vertical lines to demarcate phases

text(x=3, y=-3.5, srt=90, 'START')
text(x=27, y=-3.5, srt=90,'RAMP')
text(x=51, y=-3.5, srt=90,'STAY')
text(x=99, y=-3.5, srt=90,'END')

legend(50, -1, legend=c("SPRAAKLAB", "Traditional laboratory"), col = c(blue, yellow), lty=1:1, cex=1)

# Plot F2
plot_smooth(location_model, view="Trial", cond=list(isSpraaklabF2 = 0, FormantType='F2'), rug = F, main = "F2 over trials", ylab = "F2 (z-transformed)", xlab = "Trials",xlim=c(1,114), ylim=c(-4,3), col=c(yellow), cex.lab=1.3, cex.axis = 1.3, hide.label = TRUE)
plot_smooth(location_model, view="Trial", cond=list(isSpraaklabF2 = 1, FormantType='F2'), rug = F, main = "F2 over trials", ylab = "F2 (z-transformed)", xlab = "Trials",xlim=c(1,114), ylim=c(-4,3), col=c(blue), add = T)
par(abline(h=0, v=c(1, 25,49,97, 114), lty=3))

text(x=3, y=-3.4, srt=90,'START')
text(x=27, y=-3.4,srt=90, 'RAMP')
text(x=51, y=-3.4, srt=90,'STAY')
text(x=99, y=-3.4,srt=90, 'END')

legend(50, 3, legend=c("SPRAAKLAB", "Traditional laboratory"), col = c(blue, yellow), lty=1:1, cex=1.0)

```

# Package information 

```{r}
packages <- c("mgcv", "dplyr", "tidyr", "itsadug", "ggplot2", "reshape", "car", "nlme")

for (package_name in packages) {
  print(paste(package_name, packageVersion(package_name)))
}

R.Version()
```

